Exploratory Analysis

### 2023-11-30 Code/Draft1Rmd.Rmd
did  update

library(tidyverse)
library(ggplot2)
library(Synth)
source("Functions.R")

@@

``` {r warnings = FALSE, fig.height=5, fig.width= 10 }
# Show deaths per-capita for each state
data2 <- health_ranking |>
  mutate(od_deaths_percapita = combined_overdose_deaths / population)

@@

``` {r}
# Show deaths per-capita for each state
data2 <- health_ranking |>
  mutate(od_deaths_percapita = combined_overdose_deaths / population)

# All states and overdose death per-capita for years 2016, 2020, 2023
data2 |>
  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
  scale_x_continuous(
    trans = "log10",
    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
  ) +
  geom_density(alpha = 0.3, position = "stack") +
  facet_grid(year ~ .) +
  labs(
    title = "Density Plot of Overdose Deaths Per Capita by State",
    x = "Overdose Deaths Per Capita",
    y = "Count of Overdose Deaths",
    fill = "State"
  )

```{r}
#Create individual data sets for Oregon, Washington, and California
or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
ca_health_ranking <- health_ranking[health_ranking$state == "California",]
wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]

@@

ggplot() +
  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
  geom_point(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon"), size = 3) +
  geom_point(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California"), size = 3) +
  geom_point(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington"), size = 3) +
  geom_vline(xintercept = as.numeric("2021"), linetype = "dashed", color = "black", size = 1) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
  labs(y = "Log Combined Overdose Deaths") +
  theme_bw()
```
```

Comparing the rate of drug mortality by county depicts increases in many counties in all three states we are examining.

@@

# Oregon
health_ranking_OR %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Oregon",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

# Washington
health_ranking_WA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Washington",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()
```

```{r CA Cleveland dot plots, fig.height=7, fig.width=8}
# California 
# Save the ggplot to a variable

health_ranking_CA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in California",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

@@

## Methods

## Difference-in-Difference 
We used a difference-in-difference regression with fixed effects to investigate the impact of drug decriminalization starting in 2021 in Oregon counties compared to other states. The treatment effect shows a 80.39 decrease in overdose deaths compared to the control group and is statistically significant (p < 0.001). The fixed effects for years 2019 onwards are statistically signifcant, which indicate a time trend.

@@

### Synthetic Control 
A synthetic control analysis was conducted using Washington and Oregon as donor states. This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.

@@

### Results

Due to the limitations surrounding our synthetic control analysis, our synthetic Oregon was higher than the actual Oregon pre treatment which should not have happened. This may be due to a lack of accurate covariates used. Our analysis shows synthetic Oregon increasing more than actual Oregon, however limitations have to be noted surrounding this conclusion.

Diff in Diff:

Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.

### Limitations 
Oregon drug decriminilization has recently went into effect. The data for overdose deaths after the law has been enacted is limited with only 2021 and 2022. 
It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths. Since multiple factors can contribute to drug-related outcomes. There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health. There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon drug policies and may not be generalized to other regions or policies. There is also limited data post treatment period (2 years). Additionally, the synthetic Oregon may not be a good fit.

>>>>

library(tidyverse)
library(ggplot2)
library(Synth)
library(knitr)
source("Functions.R")

@@

``` {r warnings = FALSE, fig.height=6, fig.width= 10 }
# Show deaths per-capita for each state
data2 <- health_ranking |>
  mutate(od_deaths_percapita = combined_overdose_deaths / population)

@@

There is an increase in overdose deaths in all states after 2020, which is most likely due to the Covid-19 pandemic. The policy intervention was implemented in Oregon in 2021, and while the drug overdose rates are still increasing, Oregon appears to be increasing at a slightly slower rate than Washington and Oregon.

```{r fig.height=5, fig.width= 10}
#Create individual data sets for Oregon, Washington, and California
or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
ca_health_ranking <- health_ranking[health_ranking$state == "California",]
wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]

@@

ggplot() +
  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
  geom_point(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon"), size = 3) +
  geom_point(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California"), size = 3) +
  geom_point(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington"), size = 3) +
  geom_vline(xintercept = as.numeric("2021"), linetype = "dashed", color = "black", size = 1) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "green", "California" = "red", "Washington" = "blue")) +
  labs(y = "Log Combined Overdose Deaths") +
  theme_bw()
```

Comparing the rate of drug mortality by county depicts increases after 2020 in many counties with noticeable increases in 2022 for all three states we are examining.

@@

# Oregon
health_ranking_OR %>%
  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Oregon",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

# Washington
health_ranking_WA %>%
  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Washington",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()
```

```{r CA Cleveland dot plots, fig.height=7, fig.width=8}
# California

health_ranking_CA %>%
  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in California",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

@@

## Methods and Results

### Difference-in-Difference 
We used a difference-in-difference regression with year and county fixed effects to investigate the impact of drug decriminalization starting in 2021 in Oregon counties compared to Washington and California. The treatment estimates  an average effect of a 80.39 decrease in overdose deaths compared to the control group and is statistically significant (p < 0.001). The fixed effects for years 2019 onwards are statistically significant, which indicate a trend over time. County specific fixed effects are included in the model, represented by the coefficients for each county. Counties like Los Angeles and Riverside have positive coefficients, indicating higher overdose deaths compared to the baseline county.

@@

```{r results='asis'}
library(tidyverse)
library(broom)
library(huxtable)
library(jtools)
# Get coefficients, standard errors, and p-values
coefficients_to_display <- coef(summary(did_model_county))[c("(Intercept)", "treatment", "population"), c("Estimate", "Std. Error", "Pr(>|t|)")]

# Create a data frame for the coefficients
coefficients_df <- as.data.frame(coefficients_to_display)

# Round coefficients to 2 decimal places
rounded_coefficients <- round(coefficients_df, 3)

# Increase decimals for p-values
rounded_coefficients$`Pr(>|t|)` <- format(rounded_coefficients$`Pr(>|t|)`, nsmall = 6)  # Increase to desired number of decimals

# Format large numbers with commas
formatted_coefficients <- format(rounded_coefficients, big.mark = ",")

# Display the formatted coefficients using kable with additional formatting
kable(
  formatted_coefficients,
  format = "markdown",
  align = c("l", "c", "c", "c"),  # Adjust alignment as needed
  col.names = c("Variable", "Coefficient", "Std. Error", "P-value"),  # Rename column headers
  caption = "Regression Coefficients",  # Add a caption
  escape = FALSE  # Disable HTML escaping for formatting
)

### Synthetic Control 
A synthetic control analysis was conducted using Washington and California as donor states. This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.However, it does show that the synthetic "Oregon" with more overdose deaths if the policy intervention did not exist.

@@

### Limitations 
Oregon drug decriminalization has just recently gone into effect. The data for overdose deaths after the law has been enacted is limited with only 2021 and 2022. It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths. Since multiple factors can contribute to drug-related outcomes. There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health.

There is also limited data post treatment period (2 years). Difference-in-Differences (DiD) estimates, particularly with fixed effects, are context-specific and may lack generalizability beyond the studied time period and treatment group.The validity of DiD estimates relies on the assumption of parallel trends in the absence of treatment. While fixed effects enhance internal validity, they may limit generalizability to settings with different characteristics.

Similarly, the synthetic control analysis is specific to the context of Oregon drug policies and may not be generalized to other regions or policies. There is also limited data post treatment period (2 years). Further analysis could be done using all 50 states to find a model that best fits a synthetic Oregon.
